\chapter{Conclusions and Future Work}

\section{Conclusions}

This thesis investigated whether a game engine can serve as a viable high-fidelity simulation platform for \ac{AC}, narrowing the reality gap between algorithmic design and physical deployment. The research question was concrete: can the Unity engine be integrated with the Collektive aggregate programming library in a way that sustains the execution frequency required by \ac{AC} while providing the physics-accurate, 3D-embodied environment that abstract simulators deliberately sacrifice?

The answer is affirmative. The integration is technically feasible, and the architectural choices made to achieve it are empirically justified. The central finding is that \ac{FFI}-based communication, with Protocol Buffers as the shared data model, reduces inter-language latency to a level that is negligible relative to the cost of physics simulation and rendering. The benchmark demonstrated a mean end-to-end speedup of $456\times$ over socket-based communication, confirming that the \ac{FFI} bridge does not constitute a bottleneck and that the $20\,\text{Hz}$ execution frequency required by the sense-compute-act loop is sustainable at populations of up to 100 nodes on consumer hardware. The case study further validated that collective gradient ascent behaves correctly in both a controlled minimal environment and a high-fidelity 3D terrain, and that the richer environment exposes physical edge cases (friction, inertia, inclined surfaces) that abstract simulators would never surface. This is precisely the class of pre-deployment failure mode that motivates the entire research direction.

The implication for \ac{CAS} tooling is that the boundary between simulation and deployment can be moved meaningfully closer together without abandoning the programmability of aggregate frameworks. Collektivity does not replace scalable tools like Alchemist for population-level statistical validation; it occupies the complementary role of a high-fidelity pre-deployment stage, where the collective logic is stress-tested against physical reality before hardware is ever involved. The Unity Package Template developed as a prerequisite contribution ensures that this role is accessible: a researcher can instantiate a fully configured, CI-enabled simulation project in a single command, without expertise in game engine internals.

\section{Limitations}

Several limitations of the current work should be acknowledged. First, Collektivity is a single-machine simulator: Unity and the Collektive backend run as a single process on one host, and there is no support for distributing nodes across a network. This means that the integration currently validates collective \emph{logic} under physical conditions, but not collective \emph{communication} under realistic network constraints such as latency, packet loss or bandwidth asymmetry. Second, the execution model is globally synchronous: every node fires at the same tick, controlled by the \monospace{SimulationEngine}, whereas real \ac{CAS} deployments are inherently asynchronous, with each device running at its own frequency. This discrepancy limits the fidelity of failure-mode testing. Finally, the 100-node ceiling observed on integrated GPU hardware is a rendering and physics constraint rather than a limitation of the \ac{FFI} architecture, but it nonetheless bounds the population scales at which Collektivity can currently be used for pre-deployment validation on low-budget hardware.

\section{Future Work}

Several avenues for future work have been identified to further enhance the modularity, fidelity, and accessibility of the system.

\paragraph{Distribution via OpenUPM}
A primary limitation of the current distribution is that the project is not yet fully packageable; users must clone or fork the repository to access the simulator. A more robust approach would involve modularizing the framework so that it can be delivered as a standard Unity asset via OpenUPM~\cite{openupm}. This would allow for streamlined dependency management and easier integration into existing Unity projects without the overhead of maintaining a full repository fork.

\paragraph{Decoupled Sensor and Actuator Components}
The current implementation of the node class serves a dual purpose by representing the node entity while also housing the specific methods for sensing and acting. A more refined, decoupled architecture is envisioned where a simple identity component designates a GameObject as a node, while distinct components handle individual sensors and actuators independently. This would better align with Unity's component-based design philosophy and allow for more complex, heterogeneous swarms where different nodes carry different sensing capabilities.

\paragraph{Asynchronous Execution Model}
The most architecturally significant direction for future work is the transition from the current globally synchronous execution model to a fully asynchronous one. In the present design, the \monospace{SimulationEngine} acts as a central clock, advancing every node in lock-step each frame. This simplifies the \ac{FFI} contract (Unity makes one blocking call per tick and waits) but it diverges from the operational semantics of real \ac{AC} deployments, where each device fires its own sense-compute-act cycle independently and at a potentially different frequency.

Implementing a genuinely asynchronous model would require replacing the single blocking \ac{FFI} call with a per-node coroutine or lightweight thread, each managing its own round schedule. The \monospace{SimulationEngine} would become a coordinator rather than a clock, responsible for registering nodes and routing messages between them without imposing a shared tick. This is non-trivial precisely because the current \ac{FFI} contract is synchronous by design: the Unity C\# runtime blocks until the Kotlin side returns, and lifting this constraint requires either a non-blocking \ac{FFI} invocation pattern or a dedicated worker thread per node on the JVM side to avoid head-of-line blocking.

The benefits of making this transition are substantial. Heterogeneous round frequencies would allow the simulation to model devices with different computational budgets or energy profiles, a common scenario in \ac{IoT} deployments. Message loss and delivery delay could be injected naturally at the per-node communication layer rather than simulated as a post-hoc approximation. Most importantly, it would allow researchers to test whether a collective algorithm's self-stabilization properties hold not only under the idealized synchronous assumption but also under the realistic asynchronous conditions it will encounter in the field. Addressing this limitation would represent the most significant step toward closing the remaining gap between Collektivity and physical deployment.
